Temporal ordering and registration of images in studies of developmental dynamics

### 2014-07-10 image_analysis_paper/paper_figures2/figS6.tex
Added live imaging movies

\documentclass{pnastwo}
%\documentclass[10pt,twocolumn]{article}

%\usepackage[margin=1in]{geometry}
%\usepackage{graphicx}
\graphicspath{ {paper_figures2/} }

@@

\usepackage{float}
\usepackage{amssymb,amsfonts,amsmath}
\DeclareMathOperator*{\argmin}{arg\,min}

\makeatletter
\newcommand{\customlabel}[2]{%
\protected@write \@auxout {}{\string \newlabel {#1}{{#2}{}}}}
\makeatother

\renewcommand\thefigure{S\arabic{figure}}

\newcommand{\fig}[0]{Fig.}

@@

\begin{article}

\section{Materials and Methods}

\subsection{Image parameters}

The original resolution of each image is $1024 \times 1024$ pixels.
%
The images are subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
%
A scale bar for a representative image is shown in \fig~\ref{fig:ap_dv}.
%
When imaging, the laser settings are calibrated so that images have good dynamic range, while not saturating the detector.
%
The same laser settings are maintained within each experiment.
%
The intensity of each channel is integer-valued, and ranges from $0$ (no signal) to $255$ (saturated). 
%
We blur the nuclear signal using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function in MATLAB\textsuperscript{\textregistered}).
%
We also adjust the intensity of the nuclear signal in each image such that 1\% of the pixels are saturated 
(using the \texttt{imadjust} function in MATLAB), and then scale the intensity of the nuclear signal by half.

For the first and third data sets in the paper (images taken during cellularization), we only use the red and green channels (dpERK and Dl) in the vector diffusion maps computations, as the spatial location of the nuclei (the gray channel) does not change appreciably during this time window.
%
The nuclei in the corresponding figures are thus shown only for visual reference.
%
For the second data set of images (taken during gastrulation), we use all three channels (dpERK, Twi, and nuclei) in the vector diffusion maps computations.
%
For this particular data set, we multiply the intensity of the Twi signal by 1.5, as this channel was not completely saturated in the original imaging data.

\subsection{Algorithms}

We will demonstrate the algorithms for registration and temporal ordering using a synthetic data set.
%
The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main features of the different algorithms.
%
Motivated by the geometry of our images, we construct a sequence of concentration profiles defined on a ring, and rotate each ring randomly around its center; an example is shown in \fig~\ref{subfig:1d_example}.
%
Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of \fig~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
%
Each concentration profile is a noisy Gaussian (shown in \fig~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as a function of ``time".
%
We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_{100}$, the group of integers modulo $100$).
%
\fig~\ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.

\subsubsection{Angular synchronization \cite{singer2011angular}}

Let $ x_1, \dots, x_m$ denote the signals that we wish to align with respect to rotations;
each signal is a function defined on the unit circle (on the plane).
%
%Practically, it is discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
%rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$ by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
%
First assume that each signal $x_i$ is a {\it noisy} rotated copy of the underlying signal $x_{true}$
(which we are {\it not} given), such that
\begin{equation}
x_i = f(x_{true}, \theta_i) + \xi_i
\end{equation}
where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ radians, and $\xi_i$ is a (typically Gaussian) noise term.
%
Our goal is to recover $\theta_1, \dots, \theta_m$.
%
Up to noise,
\begin{equation} \label{eq:pairwise_rot}
x_i \approx f(x_j, \theta_i - \theta_j) ;
\end{equation}
note that \eqref{eq:pairwise_rot} does not require knowledge of $x_{true}$.
%
We can obtain an {\it estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$,
i.e., %$\theta_{ij} \approx \theta_i - \theta_j$, where
%
\begin{equation} \label{eq:opt_angle}
\theta_i - \theta_j \approx \theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
\end{equation}
%
Practically, the signals are discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$
by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
%
For the one-dimensional discretized profiles shown in \fig~\ref{fig:1d_demo}, we exhaustively search over all $n=100$ possible shifts of the signals to obtain the optimal angles in \eqref{eq:opt_angle}.
%
Alternatively, for continuous signals, an optimization algorithm
can be used \cite{ahuja2007template}.

Rather than work with the angles $\theta_{ij}$ directly, it is more convenient to consider the rotation matrices,
\begin{equation} \label{eq:R_theta}
R(\theta_{ij}) = \begin{bmatrix}
\cos(\theta_{ij}) & -\sin(\theta_{ij}) \\
\sin(\theta_{ij}) & \cos(\theta_{ij})
\end{bmatrix},
\end{equation}
which we can think of as operating on the points of the unit circle (on the plane) on which our signal is defined.
%
Successive rotations correspond to multiplication of the corresponding rotation matrices: $R(\alpha_1 + \alpha_2) = R(\alpha_1) R(\alpha_2)$.
%
Due to the orthogonality of rotation matrices, $R(-\alpha) = R(\alpha)^T$.

Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we write our procedure for general $d$ because we will later consider three-dimensional rotations).
%
We construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks, with the $i,j^{th}$ block of $H$, $H_{ij}$, defined as
\begin{equation} \label{eq:H_to_R}
H_{ij} = R(\theta_{ij}).
\end{equation}
%
%
Under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, $H_{ij} \approx R(\theta_i) R(\theta_j)^T$
%\begin{equation}
%H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
%\end{equation}
 and
\begin{equation} \label{eq:H_low_rank}
	H \approx
	\begin{bmatrix}
	R(\theta_1) \\
	R(\theta_2) \\
	\vdots \\
	R(\theta_m)
	\end{bmatrix}
	\begin{bmatrix}
	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
	\end{bmatrix}.
\end{equation}
%
It follows directly from \eqref{eq:H_low_rank} that the top block eigenvector of $H$ contains our best estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
%
Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$ ordered so that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$.
%
Then,
\begin{equation} \label{eq:R_hat}
\hat{R} =
\begin{bmatrix}
\hat{R}_1 \\
\hat{R}_2 \\
\vdots \\
\hat{R}_m
\end{bmatrix} =
\begin{bmatrix}
| & | & & | \\
\phi_1 & \phi_2 & \dots & \phi_d \\
| & | & & |
\end{bmatrix},
\end{equation}
where $\hat{R}_i \in \mathbb{R}^{d \times d}$ is (nearly) the estimate for $R(\theta_i)$.
%
To obtain our estimate of $R(\theta_i$), denoted $R_{i, est}$, we project $\hat{R}_i$ onto the closest orthogonal matrix,
\begin{equation} \label{eq:R_est}
R_{i, est} = U_i V_i^T,
\end{equation}
where $U_i$ and $V_i$ are the left and right singular vectors, respectively, of $\hat{R}_i$.
%
We adjust the sign of $\phi_1$ so that $det(R_{i, est}) = +1$, ensuring proper rotations 
(note that systematically incorporating improper rotations is also possible \cite{goemans1995improved, bandeira2013cheeger}).
%
We estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and register the signals by rotating signal $i$ by $-\theta_i$.
%
We note that, in our actual computations, the pairwise rotations $\theta_{ij}$ are computed in a discrete setting, then the overall
synchronization is performed in the continuum context to obtain $\theta_i$, and the results are rounded to give the closest
discrete shift.

Importantly, this formulation also considers {\it higher-order} consistency information.
%
For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
\begin{equation} \label{eq:triplet_consistency}
R(\theta_{ik}) R(\theta_{kj}) \approx R(\theta_i) R(\theta_k)^T R(\theta_k) R(\theta_j)^T = R(\theta_i) R(\theta_j)^T
\end{equation}
should also hold.
%
Note that
\begin{equation}
(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj});
\end{equation}
therefore, {\it all} information of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$ (and higher order
consistency information in its higher powers).
%
Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher-order considerations.

\subsubsection{Diffusion maps \cite{coifman2005geometric}}

Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional vector space), we want to find a coordinate transformation $y(x)$ that preserves local geometry: points that are ``close" in the original space should also be ``close" in the coordinates $y$.
%
The first step is to construct the matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.''
%
We use a diffusion kernel,
\begin{equation} \label{eq:dmaps_W}
W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right),
%W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
\end{equation}
where $d(x_i, x_j)$ is a pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
%
Points less than $\epsilon$ apart are thus considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
%
$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon^2$ to be $1/10^{th}$ of the median of the squared pairwise distances between data points for the first two data sets (and half of the median of the squared pairwise distances between data points for the third data set).

To find the coordinate $y$, we want solve the following optimization problem \cite{Belkin2003}
\begin{equation} \label{eq:dmaps_opt_problem}
\argmin_{y} \sum_{ij} W_{ij} (y(x_i) - y(x_j))^2.
\end{equation}
%
We first compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
\begin{equation} \label{eq:dmaps_A}
A = D^{-1} W.
\end{equation}
%
We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
%
Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors.
%
Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector; this is a trivial solution to \eqref{eq:dmaps_opt_problem}.
%
%In general, the next few eigenvectors $\phi_2, \dots, \phi_m$ give ``meaningful'' embedding coordinates for the data, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point.
%
The next eigenvector, $\phi_2$, is the (non-trivial) solution to \eqref{eq:dmaps_opt_problem}, so that $\phi_2(j)$, the $j^{th}$ entry of $\phi_2$, gives the ``new" coordinate for data point $x_j$ (i.e., $\phi_2(j) = y(x_j)$).
%
In our application, we have assumed that this {\em single} direction of variability, parameterized by $\phi_2$, is one-to-one with time.
%
Ordering the data by $\phi_2(j)$ will then, effectively, order them in time.
%
The procedure generalizes when the data lie on higher-dimensional manifolds (not just curves) in data space, where leading eigenvectors can give subsequent embedding coordinates for the data.
%

\subsubsection{Vector diffusion maps\cite{singer2012vector}}

In vector diffusion maps, given data points $x_1, \dots, x_m$, one first constructs the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
\begin{equation} \label{eq:vdm_S}
	S_{ij} = A_{ij} H_{ij}
\end{equation}
%
where $A_{ij} \in \mathbb{R}$ (defined in \eqref{eq:dmaps_A}) pertains to the diffusion kernel between data points, and $H_{ij} \in \mathbb{R}^{d \times d}$ (defined in \eqref{eq:H_to_R}) pertains to the pairwise alignment between data points.
%
It is important to note that distance $d(x_i, x_j)$ used in the diffusion kernel in \eqref{eq:dmaps_W} is the distance between data points {\it after} after pairwise alignment, i.e., the minimum distance between all possible shifts of the two data points (which is obtained in \eqref{eq:opt_pairwise}).
%
In the language of symmetry groups, this distance is a metric between the orbits induced by the relevant symmetry group.

One then computes the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{md}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{md}$ of $S$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$.
%
These eigenvectors contain information about {\it both} the optimal rotations (the ``synchronization" component) and the
variation of the data {\it after} the spatial symmetries have been factored out (in our case, their temporal variation).
%
Assuming that the data (after symmetries have been factored out) are relatively closely clustered, it is reasonable
to expect, as in angular synchronization, that the top (block) eigenvector of $S$ contains approximations of the optimal rotations,
which can be computed in the same way from \eqref{eq:R_est}.
%
We then expect subsequent eigenvectors to contain information about the main direction(s) of data variability modulo the geometric symmetries.
%%%
%%% somewhere we need to talk about compact groups
%%% discussion might be a good place because SCALING is obviously important and
%%% it is not compact......
%%%
%However, the eigenvectors now also contain information about the embedding coordinates for our images.
%

In general, the embedding coordinates are given by
\begin{equation} \label{eq:vdm_coord}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle,
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$,
%
If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations (up to a constant rotation) multiplied by the corresponding embedding coordinate (a scalar)
%
As the first diffusion maps coordinate is constant over the data, the first block eigenvector contains only the optimal rotations.
%
The second block eigenvector (eigenvectors $d+1$ through $2d$) contains the optimal rotations, each multiplied by their second diffusion maps coordinate.
%
We can therefore recover this diffusion maps coordinate by taking inner products of the columns of the second block eigenvector with columns of the first block eigenvector.
%%
%%YGK just as a favor to me, tell me how consistent the three values are -- 
%%
The $j^{th}$ embedding coordinate will be given by $\psi_{k,1}$, where $jd +1 < k \le (j+1)d$,
and we select $k$ such that the coordinate $\psi_{k, 1}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,1}$, where $k$ is the solution to
\begin{equation} \label{eq:first_embed_vdm}
\max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2. 
\end{equation}

\begin{table}[t]
%\caption{Outline of algorithm used to temporally order images}
\setlength{\fboxrule}{2pt}
\framebox[\columnwidth]{
\begin{minipage}{0.9\columnwidth}
{\bf Outline of algorithm used to register and order images}
\begin{enumerate}
\item Compute the alignments for each pair of images, as described in \eqref{eq:opt_pairwise}.
Retain both the optimal pairwise alignments (translations and rotations) as well as the minimum distances obtained when pairs are optimally aligned.
%
\item Compute the corresponding pairwise three-dimensional rotations from the translations and rotations, using \eqref{eq:angle_relations}, \eqref{eq:three_rot}, and \eqref{eq:total_rot}.
%
\item Select $\epsilon$ for use in the diffusion maps kernel. 
We often begin with using the median of the pairwise distances as $\epsilon$, and then adjust as needed.
\item Compute the matrix $S$ in \eqref{eq:vdm_S}. 
$A_{ij}$ is computed from \eqref{eq:dmaps_W} and \eqref{eq:dmaps_A}, using the pairwise distances from step 1, and $H_{ij} = R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$, which are computed in step 2.
%
\item Compute the eigenvalues  $\lambda_1, \lambda_2, \dots, \lambda_{3m}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{3m}$ of $S$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{3m}|$.
%
\item Estimate the optimal three-dimensional rotation for each image: stack the first three eigenvectors $\phi_1, \phi_2, \phi_3$ into a $3m \times 3$ matrix, and then divide this matrix into $m$ $3 \times 3$ blocks, $\hat{R}_1, \dots, \hat{R}_m$. Compute the estimated rotations for each of these $m$ blocks as in \eqref{eq:R_est}. If most of the rotations are improper ($det(R) = -1$), flip the sign of one of the eigenvectors.
\item Multiply each estimated rotation $R_{i, est}$ by $R_{1, est}^T$ (so that $R_{1, est}$ will become the identity matrix).
\item For each image $i$, compute the optimal translation and rotation to align the image by converting $R_{i, est}^T$ (note the transpose) to the corresponding translation and rotation, described in \eqref{eq:angle_invert}. 
\item Compute the first embedding coordinate $\psi_{k,1}(i)$ as described in \eqref{eq:vdm_coord} and \eqref{eq:first_embed_vdm}, where $4 \le k \le 6$.
\item To order the images, sort them by the values of this embedding coordinate $\psi_{k,1}(i)$.
\end{enumerate} 
\end{minipage}
}
\end{table}

\subsection{Registering images with respect to rotations and translations} \label{subsec:trans_rot_register}

Using angular synchronization or vector diffusion maps to register images with respect to rotations {\it and translations} requires some additional effort.
%
%We are not only interested in registering the one-dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two-dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\it and} translations.
%
%We first compute the translations and rotations to optimally align each pair of data points.
%
The first step (similar to the case of only rotations) is to compute the optimal alignments between pairs of images.
%
Practically, we have square images discretized as pixels (rather than continuous functions on the plane); note that the center of the
image square is {\em not} necessarily also the center of the embryo.
%
For each image pair $I_i$ and $I_j$ we compute
\begin{equation}\label{eq:opt_pairwise}
(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{
\begin{matrix}
\theta \in [0, 2\pi) \\
dx, dy \in [-\Delta, \Delta]\\
%dy \in [-\Delta, \Delta]
\end{matrix}
} \|g(I_j, \theta, dx, dy) - I_i \|^2.
\end{equation}
where $g(I_j, \theta, dx, dy)$ is image $I_j$ rotated around the center of the square by $\theta$ radians, then translated by $dx$ pixel widths in the $x$ direction, and finally translated by $dy$ pixel widths in the $y$ direction.
%
The norm, $\| \cdot \|$, is the Euclidean norm between the pixel intensities of the channels.
%
The domain of the image (a square) is not invariant to our rotations and translations; however, the pixels near the border of the square have zero intensity, and so the norm can be meaningfully computed as long as the main image does not ``move out of'' the original square.
%
%The norm \eqref{eq:opt_pairwise} involve an integral over the square, which is not left invariant by our rotations and
%translations; noting however that the pixels close to the border of the square have zero intensity allows a meaningful
%computation of this norm as long as the ``main feature" (the embryo) does not move out of the original square (no ``corner effects").
%
This also justifies the $\Delta$ limits in \eqref{eq:opt_pairwise}: $\Delta$ is chosen such that the main image is not split or separated when translating (we take $\Delta=10$, which corresponds to a 10\% shift in the image).
%; it does not make sense to shift the images so much that the embryo would ``move out" of the original square.
%
Image rotation is performed with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to estimate the pixel intensities after rotation.
%, since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image.
%
The missing pixels in the corners of the rotated image are taken to have zero intensity.
%
The vertical and horizontal translations are implemented by shifting the pixels;
we only consider shifts that correspond to an integer number of pixels to remove the need for interpolation when doing the translations.
%
The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
%
Therefore, instead of using an optimization procedure, we discretize the search space and exhaustively search to find the solution.
%
Our discrete translation steps are $\Delta/5$, and our discrete rotations steps are $\pi/10$.
%
%We discritize the search space of rotations into 20 possible rotations %($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$),
%and 11 possible translations in both . %($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$).
%
%We check all possible combinations for the rotation and translations that best align $I_j$ to $I_i$.
%
Although computationally demanding, this ``embarrassingly parallelizable" direct enumeration approach is not prohibitive here.
%
%Although this can be somewhat time intensive, it is not prohibitive for the data sets we consider, and can be trivially parallelized if necessary.
%
%The solution will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
%
%Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.

%As discussed above, for angular synchronization, the underlying symmetry group needs to have a real and orthogonal matrix representation.
%
%We therefore convert $ISO(2)$, the group of two-dimensional translations and rotations, to $SO(3)$, the group of three-dimensional rotations.
%

The synchronization formulation requires the matrices used to represent the symmetry group to be orthogonal \cite{singer2013spectral}.
%(for \eqref{eq:H_low_rank} to be satisfied).
%
However, the typical matrix representation for $ISO(2)$, the group of two-dimensional translations and rotations, does not satisfy this property.
%
Instead, we {\it approximately} represent rotations and translations in two dimensions using {\it three-dimensional rotation matrices}, by projecting the (two-dimensional) image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular}.
%
Rotation of the image corresponds to rotation around one principal axis in three dimensions, while translation of the image corresponds (approximately) to rotations around the other two principal axes (see \fig~\ref{fig:SO3_picture}). (Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$. However, the images we will be considering are already ``mostly centered", and so we are only interested in {\em small} translations that are well within the $[0, 2\pi)$ range.)
%
These rotation matrices are (by definition) orthogonal, and successive application of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.

To convert the optimal two-dimensional translations and rotation to corresponding (continuous) three-dimensional rotations,
we first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
\begin{equation} \label{eq:angle_relations}
%\begin{aligned}
%	\alpha_{ij} &= \theta_{ij} \\
%	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
%	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
%\end{aligned}
	\alpha_{ij} = \theta_{ij}, \; \; \;
	\beta_{ij} = \frac{dx_{ij}}{n} \times \eta_{proj}, \; \; \;
	\gamma_{ij} = \frac{dy_{ij}}{n} \times \eta_{proj}
\end{equation}
where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project the image.
%
We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere;
this portion of the sphere is small enough so that curvature effects are not prominent in the image.
%
We write rotations around the three principal axes, $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$, in terms of the three Euler angles
%\small
\begin{equation} \label{eq:three_rot}
\begin{aligned}
	R^x(\alpha) &= \begin{bmatrix}
	1 & 0 & 0 \\
    0 & \cos(\alpha) & -\sin(\alpha) \\
    0 & \sin(\alpha) & \cos(\alpha)
	\end{bmatrix} \\
	R^y(\beta) &= \begin{bmatrix}
	\cos(\beta) & 0 & \sin(\beta) \\
    0 & 1 & 0 \\
    -\sin(\beta) & 0 & \cos(\beta)
    \end{bmatrix} \\
	R^z(\gamma) &= \begin{bmatrix}
	\cos(\gamma) & -\sin(\gamma) & 0 \\
    \sin(\gamma) & \cos(\gamma) & 0 \\
    0 & 0 & 1
    \end{bmatrix}.
\end{aligned}
\end{equation}
%\normalsize
%
The total rotation, $R_{ij} \in SO(3) \subset \mathbb{R}^{3 \times 3}$, is
\begin{equation} \label{eq:total_rot}
	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij}).
%R(\alpha_{ij}, \beta_{ij}, \gamma_{ij}) =
%\begin{bmatrix}
%\cos(\beta) \cos(\gamma) & \cos(\gamma)\sin(\alpha)\sin(\beta)-\cos(\alpha) \sin(\gamma) & \cos(\alpha)\cos(\gamma)\sin(\beta)+ \sin(\alpha)\sin(\gamma) \\
%\cos(\beta) \sin(\gamma) & \cos(\alpha) \cos(\gamma)+\sin(\alpha) \sin(\beta) \sin(\gamma) & \cos(\alpha) \sin(\beta) \sin(\gamma)-\cos(\gamma) \sin(\alpha) \\
%-\sin(\beta) & \cos(\beta) \sin(\alpha) & \cos(\alpha) \cos(\beta))
%\end{bmatrix}
\end{equation}
%
%\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).

These rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$ are used in formulating the matrix $H$ in \eqref{eq:H_to_R} for angular synchronization, or the matrix $S$ in \eqref{eq:vdm_S} for vector diffusion maps to compute the optimal rotations (and, in the case of vector diffusion maps, deduce optimal temporal orderings) for the images.
%
The three-dimensional optimal rotations $R_{1, est}, \dots, R_{m, est}$ then give us the corresponding translations and rotations of the images.
%
We multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
%
$R_{i,est}^T$ then gives the rotation matrix which registers image $i$.
%
From a rotation matrix $R$, we compute the corresponding Euler angles $\alpha$, $\beta$, and $\gamma$ using the following relationships
\begin{equation} \label{eq:angle_invert}
\begin{aligned}
R_{1,1} & = \cos(\beta)\cos(\gamma) &
R_{2,1} & = \cos(\beta)\sin(\gamma) \\
R_{3,1} & = -\sin(\beta) &
R_{3,2} & = \sin(\alpha)\cos(\beta) \\
R_{3,3} & = \cos(\alpha)\cos(\beta)  &
\end{aligned}
\end{equation}
%
Inverting \eqref{eq:angle_relations} yields the desired approximate translations and rotation, and we round optimal translations to the nearest integer to remove the need for interpolation when translating the images.

\subsection{The eigenvalue spectrum}

We can use the distribution of the eigenvalues from (vector) diffusion maps to help deduce the dimensionality of the data.
%
For diffusion maps, the largest eigenvalue will always be 1 and correspond to the trivial (constant) eigenvector.
%
The eigenvalues corresponding to noise in the data form a semicircle distribution centered at the origin (when the noise is a symmetric random matrix whose entries are drawn from a Gaussian distribution) \cite{wigner1955characteristic, wigner1958distribution}.
%
We expect the eigenvalues corresponding to descriptive/meaningful embedding coordinates to be positive and ``visibly removed" from the semicircle.
%
Counting such eigenvalues which are removed from the semicircle (excluding the trivial $\lambda=1$ eigenvalue) can help obtain an estimate of the dimensionality of the data (modulo harmonics in the same dimension \cite{ferguson2010systematic}).

For vector diffusion maps, the same structure holds: we expect trivial eigenvalues, a semicircle corresponding to noise, and positive eigenvalues removed from the noise that correspond to meaningful embedding coordinates.
%
However, in vector diffusion maps, each coordinate gives rise to $d$ eigenvalues with (approximately) the same value (as each block eigenvector corresponds to $d$ eigenvalues).
%
We must then establish how many sets of $d$ eigenvalues are removed from the noise semicircle (excluding the trivial set of $d$ eigenvalues, all with  $\lambda=1$), to explore the dimensionality of the data.

\subsection{Smooth trajectories from registered and ordered images}

Once we have registered and ordered the images, we can smooth the resulting trajectory to obtain a ``stereotypic" developmental trajectory.
%
Let $x_1, \dots, x_m$ denote the set of {\em registered and ordered} images (so $x_i$ is the $i^{th}$ image in the ordered set).
%
We define the average image at time $\tau$, denoted $\overline{x}_{\tau}$ (where $1 \le \tau \le m$), as a (Gaussian) weighted average of the images,
\begin{equation}
\overline{x}_{\tau} =  \frac{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) x_i}{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) }
\end{equation}
where $\sigma$ is the scale of the Gaussian filter.
%
For the images in \fig~4, we take $\sigma^2 = 20$.
%
See \cite{kemelmacher2011exploring} for a more detailed discussion.

\bibliographystyle{pnas2}
\bibliography{background_reading/references,../../references/references}

\end{article}

\begin{figure}
\includegraphics[width=8.4cm]{ap_dv}
\caption{(Left) A lateral view of a {\it Drosophila} embryo stained with DAPI (gray), dpERK (red), and Dorsal (green). The embryo is presented so that the anterior (A) side is to the left and the posterior (P) side is to the right. The arrow indicates the position(s) where the cross-section of an embryo is imaged. (Right) A dorsoventral view of the cross-section of the {\it Drosophila} embryo. The dorsal (D) side is up and the ventral (V) side is down. For the first and third data sets (cellularization), images were collected at the focal plane $\sim 18\%$ from either the anterior or posterior pole of an embryo (arrows in the left image). During cellularization, it is known that the developmental dynamics are identical at these two regions \cite{...}. In the second data set (gastrulation), images were collected at the focal plane $\sim 18\%$ from the posterior pole of an embryo (right arrow in the left image). }
\label{fig:ap_dv}
\end{figure}

\begin{figure}
\includegraphics[width=8.4cm, trim=0cm 5.6cm 0cm 0cm, clip]{SI_fig6}
\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different times during development. The membrane length is indicated by the red bars. {\it (B)} Membrane length as a function of time. The points corresponding to the images in {\it A} are indicated with arrows.}
\label{fig:membrane_compare}
\end{figure}

>>>>

\documentclass[12pt]{minimal}

\usepackage{graphicx}

@@

\usepackage[margin=0in, paperwidth=17.8cm, paperheight=5.3cm]{geometry}

@@

\centering
%\raisebox{4.5cm}{\sffamily{A}}
%\includegraphics[width=5cm]{data1_evals}
\begin{tikzpicture}
	\node at (0,0) {\includegraphics[width=5cm]{data1_evals_dist}};
	\node at (-2.5cm, 2cm) {\sffamily{A}};
%	\draw[<->, ultra thick, cyan] (-1.4cm, -1.9cm) -- (1.7cm, -1.9cm);
	\draw[<->, ultra thick, cyan] (-1.4cm, -1.85cm) -- (1.7cm, -1.85cm);
	\draw[->, ultra thick, red] (1.97cm, -0.5cm) -- (1.97cm, -1cm);
\end{tikzpicture}
%\raisebox{4.5cm}{\sffamily{B}}
%\includegraphics[width=5cm]{data2_evals}
\begin{tikzpicture}
	\node at (0,0) {\includegraphics[width=5cm]{data2_evals_dist}};
	\node at (-2.5cm, 2cm) {\sffamily{B}};
	\draw[<->, ultra thick, cyan] (-1.2cm, -1.85cm) -- (1.5cm, -1.85cm);
	\draw[->, ultra thick, red] (1.97, -0.85cm) -- (1.97cm, -1.34cm);
\end{tikzpicture}
%\raisebox{4.5cm}{\sffamily{C}}
%\includegraphics[width=5cm]{data3_evals}
\begin{tikzpicture}
	\node at (0,0) {\includegraphics[width=5cm]{data3_evals_dist}};
	\node at (-2.5cm, 2cm) {\sffamily{C}};
	\draw[<->, ultra thick, cyan] (-0.25cm, -1.85cm) -- (0.35cm, -1.85cm);
	\draw[->, ultra thick, red] (1.97, -1.1cm) -- (1.97cm, -1.6cm);
	\draw[->, ultra thick, red] (1.52, -1.1cm) -- (1.52cm, -1.6cm);
	\draw[->, ultra thick, red] (0.64, -1.1cm) -- (0.64cm, -1.6cm);
\end{tikzpicture}
